### 2012-11-18 paper/isca40format/Introduction.tex
will be removing evict req

In this paper we will consider systems where caches form a tree hierarchy and
caches are \emph{inclusive}, that is, if a cache contains address $a$ then its
parent is guaranteed to contain address $a$, though the parent may have stale
data for $a$. Every cache or node in the system has an associated directory
which contains information about its children caches. The coherence state
changes are orchestrated by passing messages in the form of \emph{upgrade} or
\emph{downgrade} requests and responses between the caches via the directory.
Unlike bus-based protocols which lock the bus till all the required state
transitions are done thereby ensuring atomicity of state transitions, in a
message-passing protocol, requests and responses are split leading to transient
states. This creates scenarios not present in bus-based protocols, which can
lead to coherence violations or deadlocks, if the protocol is not designed or
implemented carefully.

@@

\begin{enumerate}

\item Lots of concurrent events must be dealt with without a central authority
to impose a sequence on events;

\item The state on which transitions are based is physically distributed in
space and cannot be accessed atomically;

\item Physical network may block the transmission of a message because of lack
of communication buffers or head-of-the-line blocking;

\item For each message type, determining which cache and directory states are
possible (and therefore must be dealt) with is difficult. An incomplete set of
transition rules may cause the protocol to deadlock.

\end{enumerate}

As examples of problems in designing correct protocols, we describe several
concrete and familiar scenarios from two-level systems (a private L1 cache for
each processor and a shared L2 cache):

\noindent \emph{Example 1:} A cache with a read permission for a line gets a
store request from its processor and hence, requires a write permission for
that line. The cache would send an upgrade request to the directory controller
and may eventually get a response from the directory controller. The response
would signify that the states in the other caches \emph{were} compatible with
this cache having write permission for the line. That is, the other caches
neither \emph{had} read nor write permissions for that line. But the messages
take time to propagate and the response message is necessarily based on the old
state bits (permissions) that the other caches had, not the permissions that
the other caches have at the moment the original cache gets the response. If
another cache gets a read or write permission for that line during the time the
messages are in flight, coherence would be violated.

\noindent \emph{Example 2:} A cache, already having read permission for a line,
gets a store request, and sends an upgrade request for write permission to the
directory controller. Instead of getting a response, the cache may receive a
request from the directory controller asking this cache to give up the read
permission for the line \ie an invalidate request. This could happen if the
directory controller received a request for write permission from another cache
earlier. It is not easy to determine the correct behavior of the cache
receiving the invalidate request -- whether it should ignore the invalidate
request and wait till it receives a response from the directory controller for
its request, or if it should invalidate its line. If it ignores the invalidate
request, is there any guarantee that it would receive a response for its
request for write permission, or would it deadlock?

@@

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a message-passing, directory-based setting.
Even for a 2-level cache hierarchy, designing and implementing the protocol is
quite difficult. This problem is greatly exacerbated in a multi-level cache
hierarchy, which is becoming the norm because of the increasing core counts.

In this paper, we present a substantially easier method to design
cache-coherence protocols for such systems. The method essentially breaks down
the design problem in two parts: a part that is an abstract protocol shared by
a number of concrete protocols, and another part that is specific to the
protocol to be designed and implemented. The abstract protocol is parameterized
by the number of cache levels and the number of caches in each level of the
hierarchy as well as by a predicate which involves a local ``compatibility of
state'' test. We illustrate our methodology by designing multiple
hierarchical cache-coherence protocols, notably MSI, MESI, MOSI, MOESI, both in
the presence as well as in the absence of point-to-point FIFO networks.

>>>>

In this paper we will consider systems where caches form a tree hierarchy and
caches are \emph{inclusive}, that is, if a cache contains address $a$ then its
parent is guaranteed to contain address $a$, though the parent may have stale
data for $a$. Every cache or node in the system has an associated directory
which contains information about its children caches. The coherence state
changes are orchestrated by passing requests and responses between the caches
via the directory.  Unlike bus-based protocols which lock the bus till all the
required state transitions are done thereby ensuring atomicity of state
transitions, in a message-passing protocol, requests and responses are split
leading to transient states. This creates scenarios not present in bus-based
protocols, which can lead to coherence violations or deadlocks, if the protocol
is not designed or implemented carefully.

@@

\begin{enumerate}
\item Lots of concurrent events must be dealt with without a central authority
to impose a sequence on events;
\item The state on which transitions are based is physically distributed in
space and cannot be accessed atomically;
\item Physical network may block the transmission of a message because of lack
of communication buffers or head-of-the-line blocking; 
\item The cache or directory may reach a state which can not handle a
particular message. An incomplete set of transition rules may cause the
protocol to deadlock.
\end{enumerate}

As examples of problems in designing correct protocols, we describe several
concrete and familiar scenarios from two-level systems (a private L1 cache for
each processor and a memory with a directory):

\noindent \emph{Example 1:} A cache in state $S$ for a line gets a store request
from the processor and sends a request for changing the line's state to $M$ to
the directory controller. The controller may eventually send a response
permitting the cache to go to $M$ state for that line. The response would
signify that the states in the other caches for that line \emph{were} $I$. But
the messages take time to propagate and the response is necessarily based on the
old states of other caches -- the current state of another cache may not be $I$,
violating the Single-writer-at-a-time invariant and hence breaking cache
coherence.

\noindent \emph{Example 2:} A cache in state $S$ for a line gets a store request
from the processor and sends a request for changing the line's state to $M$ to
the directory controller. Instead of getting a response permitting it to go to
$M$ state, the cache may receive a request from the directory controller asking
this cache to go to $I$ state (invalidate request)-- this could have happened if
the directory is serving some other cache's request to go to $M$ state for the
same line. It is not easy to determine the correct behavior of the former cache
on receiving the invalidate request -- whether it should ignore the invalidate
request and wait for a response to its request or whether it should invalidate
the line. Would ignoring the invalidate request lead to a deadlock?

@@

These examples give a glimpse of the complexity of designing and implementing a
correct cache coherence protocol in a directory-based non-bus setting.  Even for
a 2-level cache hierarchy, designing and implementing the protocol is quite
difficult. This problem is greatly exacerbated in a multi-level cache hierarchy,
which is becoming the norm because of the increasing core counts.

In this paper, we present a substantially easier method to design
directory-based cache-coherence protocols for such systems. The method breaks
down the design problem in two parts: a part that is shared by a number of
concrete protocols which we call \glob{}s, and another part that is specific to
the protocol to be designed and implemented, called \policy{}. \glob{}s provide
guarantees such as every request will eventually get a response, the
directory's notion of the cache states is ``conservative'' \ie a cache can not
have more permissions for a line than what the directory thinks it has, \etc;
\glob{} enables \policy{} to focus on the high level properties such as the
set of actions a cache or a directory should perform to maintain the
Single-writer-at-a-time and the Read-from-last-writer invariants.

The \glob{}s can in turn be guaranteed using a set of (\policy{} independent)
\local{}s which dictate local properties of a node such as when a request can be
handled, what state changes happen on sending a response, properties about the network,
\etc. We have formally proved that \local{}s ensure \glob{}s using the
Coq Theorem Prover \cite{}, however, a presentation of the formal proof is
beyond the scope of this paper. The \glob{}s and the \local{}s form the
cornerstone of our framework. The separation of \glob{}s and \local{}s is akin
to having an abstraction layer which hides low level details.

The \glob{}s of our framework can be used to design a wide range of cache
coherence protocols, notably MSI, MESI, MOSI, MOESI, as well as their
hierarchical version. We will show how these are designed as \policy{}s,
focusing on high-level details.

### 2012-11-18 paper/isca40format/Msi.tex
will be removing evict req

\section{Informal description of the proof}
\label{sec:informal-proof}

\newcommand{\msg}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
\newcommand{\Req}[4]{\msg{Req}{#1}{#2}{#3}{#4}}
\newcommand{\Resp}[4]{\msg{Resp}{#1}{#2}{#3}{#4}}
\newcommand{\Data}[4]{\msg{Data}{#1}{#2}{#3}{#4}}
\newcommand{\Mesg}[4]{\msg{Mesg}{#1}{#2}{#3}{#4}}
\newcommand{\FirstInv}{\emph{FirstInv}}
%\input{Atomic}

Let us walk through our design methodology for our message-passing hierarchical
directory based protocols.

\subsection{Message-passing directory-based MSI protocol for 2-levels, no
replacement evictions}
For the sake of explanation, let us start with an MSI protocol, with just two
levels of the memory hierarchy -- the L1 caches and the main memory. We will
eventually build realistic protocols with various optimizations.

Let us assume that the caches have enough capacity so that no line gets evicted
for the sake of replacement. Lines may however get evicted to enforce the
single-writer property for cache coherence, \ie when one L1 cache has
permission to modify the data for an address, no other cache has read or write
permission for that address.

An L1 cache stores the information given in Figure \ref{table:lineinfo}. The
coherence state for an address denotes the permission that a cache has for the
address: $M = \{R, W\}, S = \{R\}, I = \{\}$. State $I$ is akin to the cache not
having the data corresponding to $a$, since it can neither read or write that
cache line.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$state[c][a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for
address $a$\\
$data[c][a]$ & data in cache $c$ for address $a$\\
\hline
\end{tabularx}
\caption{Values stored by a cache for an address}
\label{table:lineinfo}
\end{figure}

In order to serve a request from the processor, if a cache has to \emph{upgrade}
its state for address $a$, \ie change from $x$ to $y$ where $x > y$
(Figure \ref{table:fns}), it can do so only after ensuring that the new state is
\emph{compatible} with the states of other caches for that address, \ie other
caches are in a state less than or equal to $toCompatible(y)$
(Figure \ref{table:fns}). This is needed to ensure the single-writer property of
the MSI protocol -- if a cache's state is $M$ for an address $a$, every other cache
must be in $I$ for address $a$. But, the caveat is that the L1 cache can access
only its own state and data, as opposed to that of other caches.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$toCompatible(s)$ & If the state of one cache is $s$ for address $a$, the
highest state of another cache for the same $a$
$\newline
toCompatible(M) = I,\newline
toCompatible(S) = S,\newline
toCompatible(I) = M
$\\
\hline
$x > y$ & Denotes if state $x$ has more permissions than state $y$.
$M > S, M > I, S > I$. Relations $\le, >, \ge$ are also defined appropriately\\
\hline
\end{tabularx}
\caption{Abstraction functions}
\label{table:fns}
\end{figure}

The memory controller has access to the data from the memory but not to the
state and data of the L1 caches. Instead, it only has access to a directory
which has a version of the coherence states of the caches for each address. The
version of the state in the directory may not be up to date, \ie may not be the
same as the one in the caches, all the time. Figure \ref{table:dirinfo} shows
the information stored in the memory and directory. We will use the word memory
and directory interchangeably for the 2-level protocol. The caches can not
communicate with each other but can communicate with the directory via messages.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$dir[c][a]$ & directory's version of the coherence state of cache $c$ for address $a$\\
$memory[a]$ & data in the memory address $a$\\
\hline
\end{tabularx}
\caption{Values stored by a directory and memory for an address}
\label{table:dirinfo}
\end{figure}

Let's us assume that we have a system which guarantees the invariants given
in Figure \ref{FirstInv}.
\floatstyle{boxed}
\restylefloat{figure}

\begin{figure}
\begin{enumerate}
\item The cache can upgrade $state[c][a]$ from $x$ to $y$ where $y > x$, \ie
\emph{downgrade} its directory state only after it receives the appropriate
authorization from the directory for address $a$. \label{cwaitresp}
\item If the cache wants to upgrade $state[c][a]$ to $x$, there is a
message \Req{c}{Dir}{a}{x} which cache $c$ can send to the directory requesting
the upgrade authorization, and the cache will eventually be able to send
the request. \label{csendreq}
\item If the directory receives an upgrade request \Req{c}{Dir}{a}{x} from cache
$c$ for address $a$, then just before receiving the request, $dir[c][a] < x$.
\label{drecvreq}
\item The directory can change $dir[c][a]$ from $x$ to $y$ where $y < x$, \ie
\emph{downgrade} its directory state only after it receives the appropriate
authorization from cache $c$ for address $a$. \label{dwaitresp}
\item If the directory wants to downgrade $dir[c][a]$ to $x$, there is a
message \Req{Dir}{c}{a}{x} which the directory can send to cache $c$ requesting
the downgrade authorization, and the directory will eventually be able to send
the request. \label{dsendreq}
\item If the directory has sent a downgrade request for an address $a$ to cache
$c$, it will eventually get the downgrade authorization response for address $a$
from cache $c$ to downgrade to $x$, where $x$ is less than or equal to the state
to which the directory sent a request to downgrade to. \label{drecvresp}
\item If $dir[c][a] = M$ just before receiving a downgrade authorization from
cache $c$ for address $a$, then the cache would have sent the data along with
the authorization. \label{drecvdata}
\item If the directory wants to send an authorization message to upgrade the
state of address $a$ in cache $c$ to $x$ (\Resp{Dir}{c}{a}{x}), it will
eventually be able to send it. \label{dsendresp}
\item The directory can send the data from the memory along with the upgrade
authorization message if it wants to. \label{dsenddata}
\item The directory's version of the cache's state is always conservative, \ie
$\forall c, a,\; dir[c][a] \ge state[c][a]$. \label{conservative}
\end{enumerate}
\caption{Initial set of invariants \FirstInv}
\label{FirstInv}
\end{figure}

\floatstyle{plain}
\restylefloat{figure}

We will walk through how the MSI coherence protocol (with no replacement
evictions) can be implemented using the above invariants and argue why it is a
correct implementation.

The directory's role is to ensure the single-writer property and that the last
updated data is being read. In order to implement the cache coherence protocols
using the invariants of Figure \ref{FirstInv}, the directory's ensures that at
all times, if for some $c, dir[c][a] = M$, the for every other $c', dir[c'][a] =
I$.

If a cache gets a request from the core which can not be handled because of lack
of permissions, it (eventually) sends an upgrade request to the directory.
\FirstInv.\ref{csendreq} ensures that such a request can eventually be sent.

If the directory receives a request from cache $c$ to upgrade the state of
address $a$ to $x$, because of \FirstInv.\ref{drecvreq}, $dir[c][a] < x$ and
since $cache[c][a] \le dir[c][a]$ because of \FirstInv.\ref{conservative},
$cache[c][a] < x$ and hence the directory must send an upgrade authorization
response. The directory sends a request \Req{Dir}{c'}{a}{toCompatible(x)} to
every cache $c'$ for which $dir[c'][a] > toCompatible(x)$. Such a request can be
sent eventually because of \FirstInv.\ref{dsendreq}.

Because of \FirstInv.\ref{drecvresp}, the directory will eventually receive
responses from each of the requested caches $c'$
(\Resp{c'}{Dir}{a}{toCompatible(x)}) authorizing the downgrade of $dir[c'][a]$
to $toCompatible(x)$. If $dir[c'][a] = M$ just before receiving the
authorization, then by \FirstInv.\ref{drecvdata}, the directory would also
receive data from $c'$. Because the directory maintains the single-writer
invariant, every other cache $c''$ would have $dir[c''][a] = I$ just before
receiving the authorization. This implies for every other cache $c''$,
$state[c''][a] = I$ by the \FirstInv.\ref{conservative} invariant, and hence
$c'$ has the most up-to-date copy for address $a$. Thus the memory is updated
with the received up-to-date copy for address $a$.

Once all the authorizations have been received, for every cache $c'$,
$dir[c'][a] \le toCompatible(x)$ (other than the one that made the request $c$).
Since the \FirstInv.\ref{conservative} invariant ensures that $cache[c'][a] \le
dir[c'][a]$, we have $cache[c'][a] \le toCompatible(x)$, ensuring the
single-writer property (because $toCompatible(M) = I$). The directory can now
change $dir[c][a]$ to $x$ and send \Resp{Dir}{c}{a}{x} authorizing $c$ to
upgrade state of address $a$ to $x$ (\FirstInv.\ref{dsendresp} ensures that such
a message can be sent). Just before sending the authorization, if $dir[c][a] =
I$, it implies that $state[c][a] = I$ because of the
\FirstInv.\ref{conservative} invariant.  Thus the cache doesn't have data
corresponding to address $a$ and the directory has to supply the most up to date
version of the address (from memory, since the memory is up to date), which it
can because of \FirstInv.\ref{dsenddata}.

The \FirstInv\ invariants essentially makes it straight forward to design the
behavior of the directory . For the sake of argument, let
\FirstInv.\ref{drecvreq} not be guaranteed. Consider a scenario the cache $c$
which has sent a request \Req{c}{Dir}{a}{M} earlier but hasn't received the
upgrade authorization response yet. Meanwhile, $c$ sends another upgrade request
\Req{c}{Dir}{a}{S}. Let us assume that the directory processed the requests in
the order in which they were sent. Once the directory has finished processing
the first request, it changes its directory state $dir[c][a]$ to $M$. When it
receives the second request, it is not easy to decide if the directory has to
drop the request or to send another response authorizing the cache to change to
$S$.

The difficulty in message passing protocols comes from the fact that it is
difficult to guarantee the above invariants, even for a 2-level memory hierarchy
discussed above as shown in the examples below.

\subsubsection*{Example 1: Inability to send messages because of network
congestion:}
For instance, the ability of a cache or the directory to send messages
(\FirstInv.\ref{dsendreq}, \FirstInv.\ref{dsendresp}, \FirstInv.\ref{csendreq},
\etc) depends upon the availability of free buffers in the network; if the
network remains congested all the time, then it would not be possible to send the
messages, and the system would deadlock.

\subsubsection*{Example 2: Multiple upgrade requests from the cache for the same
address:}
If a cache is allowed to make multiple requests to upgrade the state of an
address (say to $M$ and then to $S$) before receiving the upgrade
authorizations, then it might violate \FirstInv.\ref{drecvreq} if the requests
were processed by the directory in the order they were received.

\subsubsection*{Example 3: On sending an upgrade request, the cache waits till
it gets an authorization:}
If the behavior of the caches were such that if a cache $c$ has sent an
upgrade request \Req{c}{Dir}{a}{x} to the directory, it keeps waiting for the
upgrade authorization response \Resp{Dir}{c}{a}{x} before processing any
requests from the directory for address $a$, then the cache coupled with our
simple directory implementation would deadlock. Consider a scenario with two
caches $X$ and $Y$, with both the caches in state $S$ for address $a$. Both the
caches got a store request from the processor and want to upgrade their states
to $M$, and hence send \Req{X}{Dir}{a}{M} and \Req{Y}{Dir}{a}{M}, respectively,
to the directory. Let us say the directory receives \Req{Y}{Dir}{a}{M} first,
and in order to process it, according to our implementation of the directory,
sends a downgrade request \Req{Dir}{X}{a}{I} to cache $X$. But the cache 
does not process any request from the directory for address $a$, according to
our assumption, and hence will never send the downgrade authorization
\Resp{X}{Dir}{a}{I} to the directory, violating \FirstInv.\ref{drecvresp}, and
hence causing a deadlock.

\subsubsection*{Example 4: At the directory, authorization responses are blocked
by upgrade requests:}
Consider a scenario where two caches $X$ and $Y$ are both in state $S$ for address
$a$. $Y$ got a store request from the processor and sends an upgrade request
\Req{Y}{Dir}{a}{M} to the directory. The directory receives that request, and
sends a downgrade request \Req{X}{Dir}{a}{I} to $X$. $X$ sends a downgrade
authorization \Resp{X}{Dir}{a}{I} to the directory on receiving the downgrade
request. But the directory can not receive the authorization because it is
blocked by other requests (including \Req{Y}{Dir}{a}{M}). This violates
\FirstInv.\ref{drecvresp} and hence creates a deadlock.

These examples show that if the \FirstInv\ invariants are violated, then it
either breaks cache coherence (by creating deadlocks) or it makes correctly
designing the directory very difficult.

Before we show how the \FirstInv\ invariants can be guaranteed (which requires
the description of the behavior of the L1 caches), let us consider a
slightly more complicated system where caches can evict lines for replacement.

\subsection{Message-passing directory-based MSI protocol for 2-levels, with
replacement evictions}
We again have just two levels of memory hierarchy -- the L1 caches and the main
memory. But the caches have finite capacity. If receives a request from the core
for an address $a$ and has no space to accommodate address $a$, it evicts
another address $a'$ to make room for $a$.

In the previous scenario, where the caches didn't have to evict for replacement,
a cache only had to send downgrade authorizations to the directory on receiving
a corresponding request. But in the new scenario, a cache can send a downgrade
notification to the directory on evicting an address. Such a notification is
useful in keeping the directory's version of the cache's state up to date, so we
will restrict the family of directory-based protocols that we discuss to those
where the caches send such a notification, voluntarily.

The directory handles this notification in the exact manner by which it handled
a downgrade authorization response from the cache, so the notification message
is the same as a downgrade authorization message from cache $c$ for authorizing
the downgrade of the $dir[c][a]$ to $I$ (\Resp{c}{Dir}{a}{I}).

We will now discuss a behavior of the L1 cache that would violate the
invariant that the directory's version of the states of the caches for each
address is conservative.

\subsubsection*{Example 5: Cache sends voluntary notifications
(\Resp{c}{Dir}{a}{I}) indiscriminately}
Consider a system with two caches $X$ and $Y$. Let $state[X][a] = S, state[Y][a]
= I$ and let $dir[X][a] = S, dir[Y][a] = I$. $X$ receives a store request from
the core, and hence sends an upgrade request \Req{X}{Dir}{a}{M} to the
directory. Say $X$ was a non-blocking cache, and hence receives another load
request for address $a'$ from the core before receiving an upgrade authorization
for address $a$. $X$ does not have space in the cache for $a'$, and it decides
to evict address $a$, and sends \Resp{X}{Dir}{a}{I} to the directory, changing
$state[c][a]$ to $I$.  Meanwhile, the directory receives the request
\Req{X}{Dir}{a}{M} and since $dir[Y][a] \le I$, it responds to $X$ with an
upgrade authorization \Resp{Dir}{X}{a}{M} (along with the data). Later, the
directory receives the notification \Resp{X}{Dir}{a}{I} from $X$, and downgrades
$dir[X][a]$ to $I$. Meanwhile, $X$ receives the upgrade authorization
\Resp{Dir}{X}{a}{M} and changes $state[X][a]$ to $M$. Now,
\FirstInv.\ref{conservative} is broken, since $state[X][a] = M$ and $dir[X][a] =
I$. Let $Y$ get a load request from its core, and hence sends a
\Req{Y}{Dir}{a}{S} to the directory. The directory on receiving the request,
sends an upgrade authorization to $Y$ because $dir[X][a] = I$. This violates the
single-writer property and hence coherence.

>>>>

\section{MSI protocol}
\label{sec:msi}

\newcommand{\printall}[5]{\text{$#1 \langle #2 \rightarrow #3, #4, #5 \rangle$}}
\newcommand{\Req}[4]{\printall{Req}{#1}{#2}{#3}{#4}}
\newcommand{\Resp}[4]{\printall{Resp}{#1}{#2}{#3}{#4}}
\newcommand{\Data}[5]{\printall{Data}{#1}{#2}{#3}{#4}{#5}}

\newcommand{\lett}{\textbf{let}}
\newcommand{\send}{\textbf{send}}
\newcommand{\receive}{\textbf{receive}}
\newcommand{\pop}{\textbf{pop}}
\newcommand{\remove}{\textbf{remove}}
\newcommand{\assert}{\textbf{assert}}
\newcommand{\call}{\textbf{call}}
\newcommand{\dReq}{\textsc{handleDownReq}}
\newcommand{\uReq}{\textsc{handleUpReq}}
\newcommand{\dReqL}{\textsc{sendDownReq}}
\newcommand{\uReqL}{\textsc{sendUpReq}}
\newcommand{\dResp}{\textsc{sendDownResp}}
\newcommand{\uResp}{\textsc{sendUpResp}}

Let us walk through our framework using the example of a hierarchical MSI
protocol.

\begin{figure}
\begin{tabularx}{\linewidth}{|cX|}
\hline
$c.state[a]$ & coherence state (from set $\{M, S, I\}$) of cache $c$ for address $a$\\
$c.data[a]$ & data in cache $c$ for address $a$\\
%$c.dir[n][a]$ & $c$'s notion of the coherence state of $n$ for address $a$\\
\hline
\end{tabularx}
\caption{Notations}
\label{table:lineinfo}
\end{figure}

\begin{wrapfigure}{r}{.4\linewidth}
\centering
\begin{tabular}{|c|ccc|}
\hline
& $M$ & $S$ & $I$\\
\hline
$M$ & $=$ & $>$ & $>$\\
$S$ & $<$ & $=$ & $>$\\
$I$ & $<$ & $<$ & $=$\\
\hline
\end{tabular}
\caption{$<$ and other relations for MSI states}
\label{msi<}
\end{wrapfigure}

Each cache stores the information given in Figure \ref{table:lineinfo}. The
coherence state for an address denotes the (usual) permission that a cache has
for the address: $M = \{R, W\}, S = \{R\}, I = \{\}$. We can define a $<$
relation for coherence state values based on the permissions the states
represent. This is shown in Figure \ref{msi<}. State $I$ is akin to address $a$
not present in the cache.

\begin{wrapfigure}{r}{.46\linewidth}
\begin{subfigure}{.27\linewidth}
\centering
\begin{tabular}{|ccc|}
\hline
$M$ & $\rightarrow$ & $I$\\
$S$ & $\rightarrow$ & $S$\\
$I$ & $\rightarrow$ & $I$\\
\hline
\end{tabular}
\caption{$toCompatible$}
\label{toCompat}
\end{subfigure}
~~~~~~~~~~~~~~
\begin{subfigure}{.21\linewidth}
\centering
\begin{tabular}{|c|c|}
\hline
$M$ & $\checkmark$\\
$S$ & $\times$\\
$I$ & $\times$\\
\hline
\end{tabular}
\caption{$isModified$}
\label{isModified}
\end{subfigure}
\caption{$toCompatible$ and $isModified$ mappings}
\label{fig:msimap}
\end{wrapfigure}

If a cache $c$ receives a request for an address $a$ from the processor, it can
service the request straightaway if it has enough permissions for address $a$.
Otherwise, it has to \emph{upgrade} to a state $x$, which has enough
permissions. The cache can not upgrade its state arbitrarily, it has to ensure
that such a state change does not violate the single-writer property (Invariant
\label{singleWriter}), by \emph{downgrading}, \ie lowering the states of other
caches appropriately. We define a $toCompatible(x)$ mapping (Figure
~\ref{toCompat}), which specifies the highest state a cache can be, if another
cache is in state $x$ for the same address. We will also define a
$isModified(x)$ mapping to specify if an address in state $x$ in a cache can
have data different from what's present in the memory.

\floatstyle{boxed}
\restylefloat{figure}

\begin{figure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReq}{$c , p, a, x$}
  \If {$a$ is not present in $p$}
    \State Choose address $a'$ for eviction;
    \ForAll {$c' \in p.children$}
      \State \call{} \dReq($p , c', a', I$);
    \EndFor
    \State \call{} \dResp($p , p.parent, a', I$);
    \State Replace $a'$ with $a$ (initializing $p.state[a]$ to $I$);
  \EndIf
  \ForAll {$c' \in p.children, s.t., c' \neq c$}
    \State \dReq($p , c', a, toCompatible(x)$);
  \EndFor
  \State \uReqL($p , p.parent, a, x$);
  \State \uResp($p , c, a, x$);
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Handling upgrade to $x$ request from $c$ to $p$ for address $a$}
\label{uReq}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReq}{$p , c, a, x$}
  \ForAll {$c' \in c.children$}
    \State \dReqL($c , c', a, x$);
  \EndFor
  \State \dResp($c , p, a, x$);
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Handling downgrade to $x$ request from $p$ to $c$ for address $a$}
\label{dReq}
\end{subfigure}

\caption{MSI protocol}
\label{msi}
\end{figure}

\begin{figure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReqL}{$c , p, a, x$}
  \State \call{} \uReq($c, p, a, x$);
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending upgrade to $x$ request from $c$ to $p$ for address $a$}
\label{uReq}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReq}{$p , c, a, x$}
  \State \call{} \dReq($p, c, a, x$);
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending downgrade to $x$ request from $p$ to $c$ for address $a$}
\label{dReq}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uResp}{$p , c, a, x$}
  \If {$c.state[a] < x$}
    \If {$c.state[a] = I$}
      \State $c.data[a] \gets p.data[a]$;
    \EndIf
    \State $c.state[a] \gets x$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending upgrade to $x$ response from $p$ to $c$ for address $a$}
\label{uResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dResp}{$c , p, a, x$}
  \If {$c.state[a] > x$}
    \If {$isModified(c.state[a])$}
      \State $p.data[a] \gets c.data[a]$;
    \EndIf
    \State $c.state[a] \gets x$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending downgrade to $x$ response from $c$ to $p$ for address $a$}
\label{dResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uResp}{$p , core, a, x$}
  \State $c.data[a] \gets p.data[a]$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending upgrade to $x$ response from $p$ to $core$ for address $a$}
\label{uResp}
\end{subfigure}

\caption{Atomic MSI}
\label{atomic}
\end{figure}

\begin{figure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uReqL}{$c , p, a, x$}
  \State \send{} \Req{c}{p}{a}{x};
  \State \pop{} \Resp{p}{c}{a}{z};
  \If {$c.state[a] = I$}
    \State \pop{} \Data{p}{c}{a}{d};
    \State $c.data[a] \gets d$;
  \EndIf
  \State $c.state[a] \gets z$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending upgrade to $x$ request from $c$ to $p$ for address $a$}
\label{uReqLocal}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dReqL}{$p , c, a, x$}
  \State \send{} \Req{p}{c}{a}{x};
  \State \pop{} \Resp{c}{p}{a}{z};
  \If {$isModified(p.dir[c][a])$}
    \State \pop{} \Data{c}{p}{a}{d};
    \State $p.data[a] \gets d$;
  \EndIf
  \State $p.dir[c][a] \gets z$;
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending downgrade to $x$ request from $p$ to $c$ for address $a$}
\label{dResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uResp}{$p , c, a, x$}
  \If {$p.dir[c][a] < x$}
    \If {$p.dir[c][a] = I$}
      \State \send{} \Data{p}{c}{a}{p.data[a]};
    \EndIf
    \State $p.dir[c][a] \gets x$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending upgrade to $x$ response from $p$ to $c$ for address $a$}
\label{uResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\dResp}{$c , p, a, x$}
  \If {$c.state[a] > x$}
    \If {$isModified(c.state[a])$}
      \State \send{} \Data{c}{p}{a}{c.data[a]};
    \EndIf
    \State $c.state[a] \gets x$;
  \EndIf
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending downgrade to $x$ response from $c$ to $p$ for address $a$}
\label{dResp}
\end{subfigure}

\begin{subfigure}{\linewidth}
\begin{boxedminipage}{\linewidth}
\begin{algorithmic}
\Proc {\uResp}{$p , core, a, x$}
  \State \send{} \Data{p}{c}{a}{p.data[a]};
\EndProc
\end{algorithmic}
\end{boxedminipage}
\caption{Sending upgrade to $x$ response from $p$ to $core$ for address $a$}
\label{uResp}
\end{subfigure}

\caption{Realistic MSI protocol}
\label{realistic}
\end{figure}
